Finiteness in Cubical Type Theory

### 2020-06-28 sections/finiteness-predicates.tex
finished (first draft of?) split enumerable section

As a slightly more complex example, consider the \(\AgdaDatatype{Fin}\) type
we've been using.
Remember that split enumerability is in fact the same as a split surjection from
\(\AgdaDatatype{Fin}\) (Lemma~\ref{split-enum-is-split-surj}): to show that
\(\AgdaDatatype{Fin}\) is split enumerable, then, we need only show that it has
a split surjection from itself.
We'll prove the following slightly more general statement:
\begin{lemma}
  Every type has a split surjection into itself: the identity function.
\end{lemma}
\begin{proof}
  We provide the simple proof here in Agda:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{split-surj-ident}
  \end{agdalisting}
\end{proof}
\subsection{Decidable Equality}
We have now seen some examples of types which \emph{are} split enumerable: but
what types are excluded?
Clearly infinite types like \(\AgdaDatatype{\ensuremath{\mathbb{N}}}\) aren't
allowed, but we haven't yet answered the question of non-set types.
Certainly these can be finite in some sense, but can they be split enumerable?

As our running example of a non-set type we will use the circle:
\begin{definition}[\(S^1\)] \label{circle-def}
  The circle, \(S^1\), can be represented in HoTT as a higher inductive type.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Circle.tex]{circle-def}
  \end{agdalisting}
  We will use it here as an example of a non-set type, i.e. a type for which not
  all paths are equal.
  This also means that it does not have decidable equality.
\end{definition}

Any attempt to prove that this type is split enumerable will run into trouble:
our task here, then, is to show that such a proof is impossible.
To do will remember Hedberg's theorem
\cite{hedbergCoherenceTheoremMartinLof1998}:
\begin{theorem}[Hedberg's Theorem]
  Every discrete type is a set.
\end{theorem}
With this theorem in mind, then, if we prove that every split enumerable type is
discrete we will have proven that the circle is not split enumerable, because if
it were it would be discrete and therefore a set, which we know already it is
not.

@@

Split enumerability implies decidable equality on the underlying type.
To prove this, we will make use of the following lemma, proven in the
formalisation:
\begin{definition}[Injections]
  Injective functions are more straightforward to define constructively than
  surjective ones:
  \begin{alignat}{3}
    & \text{injective}(f) &&\coloneqq \Pi (x, y : A) , f \; x \equiv f \; y \rightarrow x \equiv y \\
    & A \rightarrowtail B &&\coloneqq \Sigma(f : A \rightarrow B) , \text{injective}(f)
  \end{alignat}
\end{definition}
\begin{lemma} \label{split-surj-to-inj}
  A split-surjection from \(A\) to \(B\) implies an injection from \(B\) to
  \(A\).
  \begin{equation}
    (A \twoheadrightarrow! \; B) \rightarrow (B \rightarrowtail A)
  \end{equation}
\end{lemma}
\begin{lemma} \label{inj-discrete}
  For any type \(A\) which injects into a discrete type \(B\), \(A\) is
  discrete.
  \begin{equation}
    \frac{
      A \rightarrowtail B \; \; \; \text{Discrete}(B)
    }{
      \text{Discrete}(A)
    }
  \end{equation}
\end{lemma}

\begin{lemma} \label{discrete-surj}
  \begin{equation}
    \frac{
      A \twoheadrightarrow! \; B \; \; \; \text{Discrete}(A)
    }{
      \text{Discrete}(B) 
    }
  \end{equation}
\end{lemma}
\begin{proof}
  This proof is can be straightforwardly derived from lemmas
  \ref{split-surj-to-inj} and \ref{inj-discrete}.
\end{proof}

>>>>

As a slightly more complex example, consider the \(\AgdaDatatype{Fin}\) type
we've been using.
Remember that split enumerability is in fact the same as a split surjection from
\(\AgdaDatatype{Fin}\) (Lemma~\ref{split-enum-is-split-surj}): to show that
\(\AgdaDatatype{Fin}\) is split enumerable, then, we need only show that it has
a split surjection from itself.
We'll prove the following slightly more general statement:
\begin{lemma}
  Every type has a split surjection into itself: the identity function.
\end{lemma}
\begin{proof}
  We provide the simple proof here in Agda:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{split-surj-ident}
  \end{agdalisting}
\end{proof}
\subsection{Decidable Equality}
We have seen now some examples of types which \emph{are} split enumerable, but
we haven't yet explored the space of types which \emph{aren't}.
Clearly this includes obviously infinite types like
\(\AgdaDatatype{\ensuremath{\mathbb{N}}}\), but it's the non-set types we're
interested in for now.
Certainly these can be finite in some sense, but can they be split enumerable?

As our running example of a non-set type we will use the circle:
\begin{definition}[\(S^1\)] \label{circle-def}
  The circle, \(S^1\), can be represented in HoTT as a higher inductive type.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Circle.tex]{circle-def}
  \end{agdalisting}
\end{definition}
The presence of the \AgdaInductiveConstructor{loop} constructor is what
precludes this type from being a set.
In sets, all paths are equal: in this type, \AgdaInductiveConstructor{loop} is a
path which is not equal to the usual identity path.

Any attempt to prove that this type is split enumerable will run into trouble:
our task here, then, is to show that such a proof is impossible.
To do so we will first remember Hedberg's theorem
\cite{hedbergCoherenceTheoremMartinLof1998}:
\begin{theorem}[Hedberg's Theorem]
  Every discrete type is a set.
\end{theorem}
With this theorem in mind, then, if we prove that every split enumerable type is
discrete we will have proven that the circle is not split enumerable, because if
it were it would be discrete and therefore a set, which we know already it is
not.

@@

\begin{definition}[Injections]
  Injective functions are more straightforward to define constructively than
  surjective ones:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Injective/Base.tex]{injective}
  \end{agdalisting} \vspace{-1.5\baselineskip}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Injective/Base.tex]{injection}
  \end{agdalisting}
\end{definition}
\begin{lemma} \label{split-surj-to-inj}
  A split-surjection from \(A\) to \(B\) implies an injection from \(B\) to
  \(A\).
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{surj-to-inj}
  \end{agdalisting}
\end{lemma}
\begin{proof} \let\qed\relax

  \begin{minipage}[t]{.9\textwidth}\vspace{-\baselineskip}
    \begin{agdalisting*}
      \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{surj-to-inj-proof}
    \end{agdalisting*}
  \end{minipage}
\end{proof}
\begin{lemma} \label{inj-discrete}
  For any type \(A\) which injects into a discrete type \(B\), \(A\) is
  discrete.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Injective/Properties.tex]{inj-discrete}
  \end{agdalisting}
\end{lemma}

\begin{lemma} \label{discrete-surj}
  If a discrete type \(A\) has a split surjection into some type \(B\), \(B\) is
  discrete.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{discrete-surj}
  \end{agdalisting}
\end{lemma}
\begin{proof}
  This proof is can be straightforwardly derived from lemmas
  \ref{split-surj-to-inj} and \ref{inj-discrete}.
\end{proof}
